---
title: "report 9"
author: "Khaled Hasan"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
library(fpp2)
library(tseries)
library(urca)
library(FinTS)
```


## Data
```{r}
GLOBALTEMPERATURE = read.csv(file = "C:\\Users/ss/Desktop/Time_series_Analysis/MyGlobalTemperetures.csv")
global_temp = ts(GLOBALTEMPERATURE[,1], start = c(1850, 1), frequency = 12)
northernhemisphere_temp = ts(GLOBALTEMPERATURE[,2], start = c(1850, 1), frequency = 12)
southernhemisphere_temp = ts(GLOBALTEMPERATURE[,3], start = c(1850, 1), frequency = 12)
LattitudinalTemps = read.csv(file = "C:\\Users/ss/Desktop/Time_series_Analysis/LatittudCuttedTemperetures.csv")
northPole = ts(LattitudinalTemps[, "X3"], start = c(1850, 1), frequency = 12)
north = ts(LattitudinalTemps[, "X2"], start = c(1850, 1), frequency = 12)
trop_north = ts(LattitudinalTemps[, "X1"], start = c(1850, 1), frequency = 12)
trop_south = ts(LattitudinalTemps[, "X4"], start = c(1850, 1), frequency = 12)
south = ts(LattitudinalTemps[, "X5"], start = c(1850, 1), frequency = 12)
southPole = ts(LattitudinalTemps[, "X6"], start = c(1850, 1), frequency = 12)
```

### Plots
```{r}
library(ggplot2)
library(gridExtra)


autoplot1 <- autoplot(window(northPole, start = c(1990, 1), freq = 12), ylab = "North Pole")

autoplot2 <- autoplot(window(north, start = c(1990, 1), freq = 12), ylab = "North Mideteranian")

autoplot3 <- autoplot(window(trop_north, start = c(1990, 1), freq = 12), ylab = "above, tropical")

autoplot4 <- autoplot(window(trop_south, start = c(1990, 1), freq = 12), ylab = "below tropical")

autoplot5 <- autoplot(window(south, start = c(1990, 1), freq = 12), ylab = "southern middle strip")

autoplot6 <- autoplot(window(southPole, start = c(1990, 1), freq = 12), ylab = "south pole")


par(mfrow = c(1, 6))
plot(autoplot1)
plot(autoplot2)
plot(autoplot3)
plot(autoplot4)
plot(autoplot5)
plot(autoplot6)

layout(matrix(c(1, 2, 3, 4, 5), nrow = 1))
grid.arrange(autoplot1, autoplot2, autoplot3, autoplot4, autoplot5, autoplot6)

```


## Forecasting from 1960
```{r}
Arima_fittng <- function(timeseries, startingPoint = start(timeseries), endingPoint = end(timeseries)){
  cutted_data = window(timeseries, start = startingPoint, end = endingPoint, freq = 12)
  t = seq_along(cutted_data)
  regressors = cbind(sin(pi/6*t), cos(pi/6*t), t)
  arima_fit = auto.arima(cutted_data, xreg = regressors, approximation = FALSE, seasonal = TRUE)
  return(arima_fit)
}

global_fitting_Arimareg = Arima_fittng(global_temp, startingPoint = c(1987, 1))
global_fitting_old = Arima_fittng(global_temp, startingPoint = c(1850, 1), endingPoint = c(1960, 1))

summary(global_fitting_Arimareg)
```


```{r}
t_forecast = (0):(720)
forecasting_regressors = cbind(sin(t_forecast*pi/6), cos(t_forecast*pi/6), t_forecast)
global_fitting_Arimareg%>% forecast(xreg = forecasting_regressors) %>% autoplot()
```

An observation: There seems to be a periodicity with higher number of years.


## Different Earth strips
- northPole: latitude from $60^o$ to $90^o$.
- north: latitude from $30^o$ to $60^o$.
- 

```{r}
northPole_fitting_Arimareg = Arima_fittng(northPole, startingPoint = c(1980, 1))
north_fitting_Arimareg = Arima_fittng(north, startingPoint = c(1980, 1))
trop_north_fitting_Arimareg = Arima_fittng(trop_north, startingPoint = c(1980, 1))
trop_south_fitting_Arimareg = Arima_fittng(trop_south, startingPoint = c(1980, 1))
south_fitting_Arimareg = Arima_fittng(south, startingPoint = c(1980, 1))
southPole_fitting_Arimareg = Arima_fittng(southPole, startingPoint = c(1980, 1))
```




```{r}

linear_coef <- function(DATA, x, Ord, sOrd, radius = 2){
  
  temporary_data = window(DATA, start = c(x-radius, 1), end = c(x+radius, 1))
  
  new_t <- seq_along(temporary_data)
  
  temporary_xreg = cbind(sin(new_t*pi/6), cos(new_t*pi/6), new_t)
  
  temporary_model = arima(temporary_data, order = Ord, seasonal = sOrd, xreg = temporary_xreg)
  
  std_error <- sqrt(diag(vcov(temporary_model)))
  
  return(c(as.numeric(temporary_model$coef["new_t"]), as.numeric(sqrt(diag(vcov(temporary_model)))["new_t"])))
}


plot_p_vals <- function(timeseries, arima_fit, r){
  
  
  ord = arima_fit$arma
  
  p = ord[1]; q = ord[2]; P = ord[3]; Q = ord[4]; period = ord[5]; d = ord[6]; D = ord[7];
  
  parameters =c()
  
  errors = c()
  
  p_values = c()
  
  rad = r
  
  sp = 1850
  
  fp = 2010
  
  for (i in (sp + rad):(fp - rad)){
      u = linear_coef(timeseries,  i, c(p, d, q), c(P, D, Q) , rad)
      parameters <- cbind(parameters, u[1])
      errors <- cbind(errors, u[2])
      p = 2*pnorm(min(0, 2*u[1]), mean = u[1], sd = u[2], lower.tail = TRUE)
      p_values = cbind(p_values, p)
  }
  
  t_vals = c((sp + rad):(fp - rad))
  
  plot(t_vals, -log10(p_values), type='b', main=paste("p\nradius: ", toString(rad)))
  
}

plot_p_vals(global_temp, arima_fit = global_fitting_Arimareg, 20)
```



```{r}
plot_p_vals(northPole, arima_fit = northPole_fitting_Arimareg, 20)
```

```{r}
plot_p_vals(north, arima_fit = north_fitting_Arimareg, 20)
```


```{r}
plot_p_vals(trop_north, arima_fit = trop_north_fitting_Arimareg, 20)
```



```{r}
plot_p_vals(trop_south, arima_fit = trop_south_fitting_Arimareg, 20)
```




```{r}
plot_p_vals(south, arima_fit = south_fitting_Arimareg, 20)
```



```{r}
plot_p_vals(southPole, arima_fit = southPole_fitting_Arimareg, 20)
```
error, did not work


## plots

```{r}
library(ggplot2)
library(gridExtra)


autoplot1 <- autoplot(window(global_temp, start = c(1990, 1), freq = 12), ylab = "global")

autoplot2 <- autoplot(window(north, start = c(1990, 1), freq = 12), ylab = "Northern temps")

autoplot3 <- autoplot(window(south, start = c(1990, 1), freq = 12), ylab = "southern temps")

autoplot4 <- autoplot(window(tropical, start = c(1990, 1), freq = 12), ylab = "tropical temps")

autoplot5 <- autoplot(window(south_noPole, start = c(1990, 1), freq = 12), ylab = "southern ex. Pole temps")


par(mfrow = c(1, 5))
plot(autoplot1)
plot(autoplot2)
plot(autoplot3)
plot(autoplot4)
plot(autoplot5)

layout(matrix(c(1, 2, 3, 4, 5), nrow = 1))
grid.arrange(autoplot1, autoplot2, autoplot3, autoplot4, autoplot5)
```

There are few interesting points about the above data. First, we notice that excluding north-pole did not seem to shift the time phase, but it significantly shifted the data upwards, we also notice that the tropical data seems to agree with the northern phase, suggesting that the region between latitudes $-30 ^o \text{ and } 30 ^o$ has more mass in the northern hemisphere, but what is rather more interesting, is the seasonal patters that defies the usual sinusoidal pattern, this behavior is to be studied.


### Arima Fitting
```{r}
tropical_autoArima_fit = auto.arima(window(tropical, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, stepwise = FALSE, allowdrift = TRUE)

summary(tropical_autoArima_fit)
```


```{r}
checkresiduals(tropical_autoArima_fit)
```

doing the same with different latitudinal cuts:
```{r}
global_autoArima_fit = auto.arima(window(global_temp, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, allowdrift = TRUE)

north_autoArima_fit = auto.arima(window(north, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, allowdrift = TRUE)


south_autoArima_fit = auto.arima(window(south, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, allowdrift = TRUE)
south_np_autoArima_fit = auto.arima(window(south_noPole, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, allowdrift = TRUE)


northernhemisphereArimaFit = auto.arima(window(northernhemisphere_temp, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, allowdrift = TRUE)
southernhemisphereArimaFit = auto.arima(window(southernhemisphere_temp, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, allowdrift = TRUE)
```


```{r}
{
  print(global_autoArima_fit$arma)
  print(north_autoArima_fit$arma)
  print(south_autoArima_fit$arma)
  print(south_np_autoArima_fit$arma)
  print(northernhemisphereArimaFit$arma)
  print(southernhemisphereArimaFit$arma)
}
```

```{r}
tropical_autoArima_fit %>% forecast(h = 24) %>% autoplot()
```



```{r}
{
  print(global_autoArima_fit$aicc)
  print(north_autoArima_fit$aicc)
  print(tropical_autoArima_fit$aicc)
  print(south_autoArima_fit$aicc)
  print(south_np_autoArima_fit$aicc)
  print(northernhemisphereArimaFit$aicc)
  print(southernhemisphereArimaFit$aicc)
}
```

```{r}
Arima_fittng <- function(timeseries, startingPoint = start(timeseries), endingPoint = end(timeseries)){
  cutted_data = window(timeseries, start = startingPoint, end = endingPoint, freq = 12)
  t = seq_along(cutted_data)
  regressors = cbind(sin(pi/6*t), cos(pi/6*t), t)
  arima_fit = auto.arima(cutted_data, xreg = regressors, approximation = FALSE, stepwise = FALSE, seasonal = TRUE)
  return(arima_fit)
}

global_fitting_Arimareg = Arima_fittng(global_temp, startingPoint = c(1980, 1))

north_reg = Arima_fittng(north, startingPoint = c(1980, 1))

tropical_reg = Arima_fittng(tropical, startingPoint = c(1980, 1))

south_reg = Arima_fittng(south, startingPoint = c(1980, 1))

south_np_reg = Arima_fittng(south_noPole, startingPoint = c(1980, 1))


summary(north_reg)
```


```{r}
checkresiduals(tropical_reg)
```
Notice that the regression models here did not fit the data as well as the global case, despite using periodic fits to handle seasonality.


```{r}
forecast_reg <- function(model, h){
  t_forecast = (0):(h)
  forecasting_regressors = cbind(sin(t_forecast*pi/6), cos(t_forecast*pi/6), t_forecast)
  model %>% forecast(xreg = forecasting_regressors) %>% autoplot() %>% return()
}

forecast_reg(tropical_reg, 24)
```


### Triying finer frequencies:

```{r}
Arima_fittng2 <- function(timeseries, startingPoint = start(timeseries), endingPoint = end(timeseries)){
  cutted_data = window(timeseries, start = startingPoint, end = endingPoint, freq = 12)
  t = seq_along(cutted_data)
  regressors = cbind(sin(pi/6*t), cos(pi/6*t), sin(pi/12*t), cos(pi/12*t), t)
  arima_fit = auto.arima(cutted_data, xreg = regressors, approximation = FALSE, seasonal = TRUE)
  return(arima_fit)
}

tropical_reg2 = Arima_fittng2(tropical, startingPoint = c(1980, 1))

summary(tropical_reg2)
checkresiduals(tropical_reg2)
```


```{r}

t_forecast = (0):(24)
forecasting_regressors = cbind(sin(t_forecast*pi/6), cos(t_forecast*pi/6), sin(t_forecast*pi/12), cos(t_forecast*pi/12), t_forecast)
tropical_reg2 %>% forecast(xreg = forecasting_regressors) %>% autoplot() %>% return()

```


```{r}
Arima_fittng3 <- function(timeseries, startingPoint = start(timeseries), endingPoint = end(timeseries)){
  cutted_data = window(timeseries, start = startingPoint, end = endingPoint, freq = 12)
  t = seq_along(cutted_data)
  regressors = cbind(sin(pi/6*t), cos(pi/6*t), sin(pi/12*t), cos(pi/12*t), sin(pi/18*t), cos(pi/18*t), t)
  arima_fit = auto.arima(cutted_data, xreg = regressors, approximation = FALSE, seasonal = TRUE)
  return(arima_fit)
}

tropical_reg3 = Arima_fittng3(tropical, startingPoint = c(1980, 1))

summary(tropical_reg3)
checkresiduals(tropical_reg3)
```

## Dow Jones data:

src: https://www.investing.com/indices/us-30-historical-data

```{r}
DOWJONES = read.csv(file= "C:\\Users/ss/Desktop/Time_series_Analysis/Dow Jones Industrial Average Historical Data.csv")
dowjones = ts(DOWJONES[, 2], start=c(1970, 2), frequency = 12)
autoplot(dowjones)
```


Now I will fit it to autoArima
```{r}
dowjones_autoArima_fit = auto.arima(dowjones, approximation = FALSE, seasonal = TRUE, stepwise = FALSE, allowdrift = TRUE)

summary(dowjones_autoArima_fit)
```

try logarithmic transformation:
```{r}
Ldowjones = log10(dowjones)
autoplot(Ldowjones)
```


```{r}
Ldowjones_autoArima_fit = auto.arima(Ldowjones, approximation = FALSE, seasonal = TRUE, stepwise = FALSE, allowdrift = TRUE)


summary(Ldowjones_autoArima_fit)
```

```{r}
checkresiduals(Ldowjones_autoArima_fit)
```

```{r}
summary(auto.arima(Ldowjones, approximation = FALSE, seasonal = TRUE, stepwise = FALSE, allowdrift = FALSE))
```



