---
title: "Report11"
author: "Khaled Hasan"
date: "`r Sys.Date()`"
output: pdf_document

---

toadd: table of content

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(fpp2)
library(tseries)
library(urca)
library(FinTS)
```


# 1.Data
```{r}
GLOBALTEMPERATURE = read.csv(file = "C:\\Users/ss/Desktop/Time_series_Analysis/MyGlobalTemperetures_LandOnly.csv")
global_temp = ts(GLOBALTEMPERATURE[,1], start = c(1850, 1), frequency = 12)
northernhemisphere_temp = ts(GLOBALTEMPERATURE[,2], start = c(1850, 1), frequency = 12)
southernhemisphere_temp = ts(GLOBALTEMPERATURE[,3], start = c(1850, 1), frequency = 12)

LattitudinalTemps = read.csv(file = "C:\\Users/ss/Desktop/Time_series_Analysis/LatittudCuttedTempereturesLandOnly.csv")
northPole = ts(LattitudinalTemps[, "X3"], start = c(1850, 1), frequency = 12)
north = ts(LattitudinalTemps[, "X2"], start = c(1850, 1), frequency = 12)
trop_north = ts(LattitudinalTemps[, "X1"], start = c(1850, 1), frequency = 12)
trop_south = ts(LattitudinalTemps[, "X4"], start = c(1850, 1), frequency = 12)
south = ts(LattitudinalTemps[, "X5"], start = c(1850, 1), frequency = 12)
southPole = ts(LattitudinalTemps[, "X6"], start = c(1850, 1), frequency = 12)
```

## 1.1.Plots
```{r}
library(ggplot2)
library(gridExtra)


autoplot1 <- autoplot(window(northPole, start = c(1990, 1), freq = 12), ylab = "North Pole")

autoplot2 <- autoplot(window(north, start = c(1990, 1), freq = 12), ylab = "North Mideteranian")

autoplot3 <- autoplot(window(trop_north, start = c(1990, 1), freq = 12), ylab = "above, tropical")

autoplot4 <- autoplot(window(trop_south, start = c(1990, 1), freq = 12), ylab = "below tropical")

autoplot5 <- autoplot(window(south, start = c(1990, 1), freq = 12), ylab = "southern middle strip")

autoplot6 <- autoplot(window(southPole, start = c(1990, 1), freq = 12), ylab = "south pole")


par(mfrow = c(1, 6))
plot(autoplot1)
plot(autoplot2)
plot(autoplot3)
plot(autoplot4)
plot(autoplot5)
plot(autoplot6)

layout(matrix(c(1, 2, 3, 4, 5), nrow = 1))
grid.arrange(autoplot1, autoplot2, autoplot3, autoplot4, autoplot5, autoplot6)

```


#2. Different Earth strips
- northPole: latitude from $60^o$ to $90^o$.
- north: latitude from $30^o$ to $60^o$.
- trop_north: latitude from $0^o$ to $30^o$.
- trop_south: latitude from $0^o$ to $-30^o$.
- south: latitude from $-30^o$ to $-60^o$.
- southPole: latitude from $-60^o$ to $-90^o$.
```{r}
Arima_fittng <- function(timeseries, startingPoint = start(timeseries), endingPoint = end(timeseries)){
  cutted_data = window(timeseries, start = startingPoint, end = endingPoint, freq = 12)
  t = seq_along(cutted_data)
  regressors = cbind(sin(pi/6*t), cos(pi/6*t), t)
  arima_fit = auto.arima(cutted_data, xreg = regressors, approximation = FALSE, seasonal = TRUE)
  return(arima_fit)
}

global_fitting_Arimareg = Arima_fittng(global_temp, startingPoint = c(1985, 1))

summary(global_fitting_Arimareg)
```



```{r}
northPole_fitting_Arimareg = Arima_fittng(northPole, startingPoint = c(1980, 1))
north_fitting_Arimareg = Arima_fittng(north, startingPoint = c(1980, 1))
trop_north_fitting_Arimareg = Arima_fittng(trop_north, startingPoint = c(1980, 1))
trop_south_fitting_Arimareg = Arima_fittng(trop_south, startingPoint = c(1980, 1))
south_fitting_Arimareg = Arima_fittng(south, startingPoint = c(1980, 1))
southPole_fitting_Arimareg = Arima_fittng(southPole, startingPoint = c(1980, 1))
```


```{r}

model_details <- data.frame(c(northPole_fitting_Arimareg$arma), c(north_fitting_Arimareg$arma), c(trop_north_fitting_Arimareg$arma), c(trop_south_fitting_Arimareg$arma),c(south_fitting_Arimareg$arma), c(southPole_fitting_Arimareg$arma))

model_details
```



```{r}
summary(northPole_fitting_Arimareg)
```

```{r}
summary(north_fitting_Arimareg)
```

```{r}
summary(trop_north_fitting_Arimareg)
```

```{r}
summary(trop_south_fitting_Arimareg)
```


```{r}
summary(south_fitting_Arimareg)
```

```{r}
summary(southPole_fitting_Arimareg)
```


```{r}

# Helping functions

linear_coef <- function(DATA, x, Ord, sOrd, radius = 2){
  
  temporary_data = window(DATA, start = c(x-radius, 1), end = c(x+radius, 1))
  
  new_t <- seq_along(temporary_data)
  
  temporary_xreg = cbind(sin(new_t*pi/6), cos(new_t*pi/6), new_t)
  
  temporary_model = arima(temporary_data, order = Ord, seasonal = sOrd, xreg = temporary_xreg)
  
  std_error <- sqrt(diag(vcov(temporary_model)))
  
  return(c(as.numeric(temporary_model$coef["new_t"]), as.numeric(sqrt(diag(vcov(temporary_model)))["new_t"])))
}


plot_Global_warming <- function(timeseries, arima_fit){
  
  #arima_fit = Arima_fittng(timeseries, startingPoint = c(1980, 1))
  
  
  ord = arima_fit$arma
  
  p = ord[1]; q = ord[2]; P = ord[3]; Q = ord[4]; period = ord[5]; d = ord[6]; D = ord[7];
  
  parameters =c()
  
  errors = c()
  
  rad = 20
  
  sp = 1850
  
  fp = 2023
  
  for (i in (sp + rad):(fp - rad)){
      u = linear_coef(timeseries,  i, c(p, d, q), c(P, D, Q) , rad)
      parameters <- cbind(parameters, u[1])
      errors <- cbind(errors, u[2])
  }
  
  t_vals = c((sp + rad):(fp - rad))
  
  MyPlot = (plot(t_vals, parameters, type='b', main=paste("c\nradius: ", toString(rad)), ylim = c(-0.010, 0.010)) +
    arrows(x0=t_vals, y0=parameters-2*errors, x1 = t_vals, y1=parameters+2*errors, code=3, angle = 90, length = 0.1))+
    grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)      # Grid line width
  
  return(list(arima_fit, parameters, errors, MyPlot))
}


plot_p_vals <- function(timeseries, arima_fit, r){
  
  
  ord = arima_fit$arma
  
  p = ord[1]; q = ord[2]; P = ord[3]; Q = ord[4]; period = ord[5]; d = ord[6]; D = ord[7];
  
  parameters =c()
  
  errors = c()
  
  p_values = c()
  
  rad = r
  
  sp = 1850
  
  fp = 2010
  
  for (i in (sp + rad):(fp - rad)){
      u = linear_coef(timeseries,  i, c(p, d, q), c(P, D, Q) , rad)
      parameters <- cbind(parameters, u[1])
      errors <- cbind(errors, u[2])
      p = 2*pnorm(min(0, 2*u[1]), mean = u[1], sd = u[2], lower.tail = TRUE)
      p_values = cbind(p_values, p)
  }
  
  t_vals = c((sp + rad):(fp - rad))
  
  plot(t_vals, -log10(p_values), type='b', main=paste("p\nradius: ", toString(rad)))
  
}

plot_p_vals(global_temp, arima_fit = global_fitting_Arimareg, 20)+
    grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)      # Grid line width
  
```



```{r}
plot_p_vals(northPole, arima_fit = northPole_fitting_Arimareg, 20)
```

```{r}
plot_p_vals(north, arima_fit = north_fitting_Arimareg, 20)
```


```{r}
plot_p_vals(trop_north, arima_fit = trop_north_fitting_Arimareg, 20)
```



```{r}
plot_p_vals(trop_south, arima_fit = trop_south_fitting_Arimareg, 20)
```




```{r}
plot_p_vals(south, arima_fit = south_fitting_Arimareg, 20)
```



```{r}
# plot_p_vals(southPole, arima_fit = southPole_fitting_Arimareg, 20)
```
Error. Again.

```{r, message=FALSE, warning=FALSE, error=TRUE}
plot_Global_warming(northPole, arima_fit = northPole_fitting_Arimareg)
```

```{r, message=FALSE, warning=FALSE}
plot_Global_warming(north, arima_fit = north_fitting_Arimareg)
```

```{r, message=FALSE, warning=FALSE, error=TRUE}
# plot_Global_warming(trop_north, arima_fit = trop_north_fitting_Arimareg)
```


```{r, message=FALSE, warning=FALSE, error=TRUE}
plot_Global_warming(trop_south, arima_fit = trop_south_fitting_Arimareg)
```


```{r, message=FALSE, warning=FALSE, error=TRUE}
plot_Global_warming(south, arima_fit = south_fitting_Arimareg)
```










# Separating Months:
```{r}
MonthsSeparatedData = read.csv("C:/Users/ss/Desktop/Time_series_Analysis/NorthTemperetures_LandOnly_sepByMonth.csv")
Jans = ts(data = MonthsSeparatedData[, 1], start = c(1850), end = c(2022), frequency = 1)
Febs = ts(data = MonthsSeparatedData[, 2], start = c(1850), end = c(2022), frequency = 1)
Mars = ts(data = MonthsSeparatedData[, 3], start = c(1850), end = c(2022), frequency = 1)
Aprs = ts(data = MonthsSeparatedData[, 4], start = c(1850), end = c(2022), frequency = 1)
Mays = ts(data = MonthsSeparatedData[, 5], start = c(1850), end = c(2022), frequency = 1)
Juns = ts(data = MonthsSeparatedData[, 6], start = c(1850), end = c(2022), frequency = 1)
Juls = ts(data = MonthsSeparatedData[, 7], start = c(1850), end = c(2022), frequency = 1)
Augs = ts(data = MonthsSeparatedData[, 8], start = c(1850), end = c(2022), frequency = 1)
Seps = ts(data = MonthsSeparatedData[, 9], start = c(1850), end = c(2022), frequency = 1)
Octs = ts(data = MonthsSeparatedData[, 10], start = c(1850), end = c(2022), frequency = 1)
Novs = ts(data = MonthsSeparatedData[, 11], start = c(1850), end = c(2022), frequency = 1)
Decs = ts(data = MonthsSeparatedData[, 12], start = c(1850), end = c(2022), frequency = 1)
```


## Plotting

```{r}
library(ggplot2)
library(gridExtra)


plot1 <- autoplot(Jans)

plot2 <- autoplot(Febs)

plot3 <- autoplot(Mars)

plot4 <- autoplot(Aprs)

plot5 <- autoplot(Mays)

plot6 <- autoplot(Juns)

plot7 <-autoplot(Juls)

plot8 <- autoplot(Augs)

plot9 <- autoplot(Seps)

plot10 <- autoplot(Octs)

plot11 <- autoplot(Novs)

plot12 <- autoplot(Decs)



par(mfrow = c(2, 6))
plot(plot1)
plot(plot2)
plot(plot3)
plot(plot4)
plot(plot5)
plot(plot6)
plot(plot7)
plot(plot8)
plot(plot9)
plot(plot10)
plot(plot11)
plot(plot12)

layout(matrix(c(c(1, 2, 3, 4, 5, 6), c(1, 2, 3, 4, 5, 6)), nrow = 2))
grid.arrange(plot1, plot2, plot3, plot4, plot5, plot6, plot7, plot8, plot9, plot10, plot11, plot12)
```


```{r}
ggAcf(Juns)
```
There is no apparent seasonality in this figure.
```{r}
Pacf(diff(diff(Juns)))
```

the acf of the differenced data seems to suggest cyclic rather than seasonal behavior.


```{r}
adf.test(Juns)
```


```{r}
adf.test(diff(Augs))
```

```{r}
kpss.test(diff(Augs), null = "Level")
```



```{r}
JansArimaFit = auto.arima(Jans, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(JansArimaFit)
```

```{r}
FebsArimaFit = auto.arima(Febs, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(FebsArimaFit)
```


```{r}
MarsArimaFit = auto.arima(Mars, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(MarsArimaFit)
```

```{r}
AprsArimaFit = auto.arima(Aprs, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(AprsArimaFit)
```


```{r}
MaysArimaFit = auto.arima(Mays, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(MaysArimaFit)
```


```{r}
JunsArimaFit = auto.arima(Juns, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(JunsArimaFit)
```


```{r}
JulsArimaFit = auto.arima(Juls, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(JulsArimaFit)
```


```{r}
AugsArimaFit = auto.arima(Augs, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(AugsArimaFit)

plot(AugsArimaFit)
```


```{r}
SepsArimaFit = auto.arima(Seps, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(SepsArimaFit)
```


```{r}
OctsArimaFit = auto.arima(Octs, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(OctsArimaFit)
```


```{r}
NovsArimaFit = auto.arima(Novs, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(NovsArimaFit)
```

```{r}
DecsArimaFit = auto.arima(Decs, stepwise = FALSE, approximation = FALSE, allowdrift = TRUE)

summary(DecsArimaFit)
```


```{r}
# Load the grid package
library(grid)

# Create a basic plot with a custom grid layout
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 3)))

# Define a function to place a plot in a specific grid cell
plot_in_grid <- function(row, col, plot_function) {
  vp <- viewport(layout.pos.row = row, layout.pos.col = col)
  pushViewport(vp)
  plot_function()
  popViewport()
}
```

plotting ma1 coefficient for each month:
```{r}
ma1_coeafficients = c(JansArimaFit$coef['ma1'], FebsArimaFit$coef['ma1'], MarsArimaFit$coef['ma1'],
                      AprsArimaFit$coef['ma1'], MaysArimaFit$coef['ma1'], JunsArimaFit$coef['ma1'],
                      JulsArimaFit$coef['ma1'], AugsArimaFit$coef['ma1'], SepsArimaFit$coef['ma1'],
                      OctsArimaFit$coef['ma1'], NovsArimaFit$coef['ma1'], DecsArimaFit$coef['ma1'])
c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
plot(c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12), ma1_coeafficients, xlab="Month")
```


## Minmum window size:

Inspired from Wacyl's work.

The smaller the AIC the better the model, however,

```{r}
months_models = cbind(c(JansArimaFit), c(FebsArimaFit), c(MarsArimaFit),
                  c(AprsArimaFit), c(MaysArimaFit), c(JunsArimaFit),
                  c(JulsArimaFit), c(AugsArimaFit), c(SepsArimaFit),
                  c(OctsArimaFit), c(NovsArimaFit), c(DecsArimaFit))

plot_arima_coef <- function(models, c){
  
  model_coefficients <- c()
  errors <- c()
  

  for (model in 1:length(models[1, ])){
      model_coefficients <- cbind(model_coefficients, models[, model]$coef[c])
      std_error <- sqrt(diag(models[, model]$var.coef))[c]
      errors <- cbind(errors, std_error)
  }
  
  t_vals<-seq_along(models[1,])
  
  MyPlot = (plot(t_vals, model_coefficients, type='b', ylim = c(-2, 1)) +
    arrows(x0=t_vals, y0=model_coefficients-2*errors, x1 = t_vals, y1=model_coefficients+2*errors, code=3, angle = 90, length = 0.1))+
    grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)      # Grid line width
}

plot_arima_coef(months_models, "ma1")
```



```{r}
plot_arima_coef(months_models, "ar1")
```

```{r}
checkresiduals(AugsArimaFit)
```



