---
title: "Task 4"
author: "Khaled Hasan"
date: "2024-03-15"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
## import libraries:

```{r}
library(fpp2)
library(tseries)
library(urca)
library(TSA)
library(signal)
library(spectral)
```


# Import data

## Real interest rate

```{r}
REAINTRATREARAT10Y <- read.csv("C:\\Users\\ss\\Downloads\\REAINTRATREARAT10Y.csv")
interest <- ts(REAINTRATREARAT10Y[, "REAINTRATREARAT10Y"], frequency = 12, start = c(1982, 1))
autoplot(interest)
```


## Market Yield on U.S. Treasury

```{r}
Treasury <- read.csv("C:/Users/ss/Desktop/Time_series_Analysis/GS10.csv")
treasury <- ts(Treasury[, "GS10"], frequency = 12, start = c(1953, 4))
autoplot(treasury)
```

Now we cut windows beyond the year 2000
```{r}
interest2 <- window(interest, frequency = 12, start = c(2000, 1))
autoplot(interest2)
```


```{r}
treasury2 <- window(treasury, frequency = 12, start = c(2000, 1))
autoplot(treasury2)
```


# Analysis of the Treasury dataset

### Checking the seasonality:


```{r}
ggseasonplot(treasury2)
```

```{r}
ggseasonplot(treasury2, polar = TRUE)
```

```{r}
ggsubseriesplot(treasury2)
```


```{r}
ggAcf(treasury2)
```

The data does not show clear seasonality.
```{r}
periodogram(treasury2)
```

```{r}
periodogram(interest2)
```


### Checking Stationarity

The KPSS test ($H_0:$ the timeseries is stationary)

```{r}
kpss.test(treasury2)
```

$p<0.05$ therefore, $H_0$ is rejected, and the time-series is not stationary.


The Dicky-Fuller test ($H_0:$ there exist a unit root which implies a non-stationary time series):
```{r}
adf.test(treasury2)
```
$p$ value is large, and therefore, I will assume a non-stationary data

```{r}
ndiffs(treasury2)
```
The (ndiffs) function in R implies that one differentiation is necessary to reach stationarity.

```{r}
checkresiduals(diff(treasury2))
```

The L-jung box test does not correspond to a white noise.



# Arima models:

### Auto Arima
```{r}
autofit = auto.arima(treasury2, approximation = FALSE, stepwise = FALSE)
summary(autofit)
```

Checking residuals of the automated model:
```{r}
checkresiduals(autofit)
```

since the $p$ value is large, the null hypothesis $(H_0)$ is not rejected, and the residuals are not correlated.

```{r}
ggAcf(diff(treasury2))
```

```{r}
ggPacf(diff(treasury2))
```

The models suggested by ACF and PACF are $(0, 1, 1)$, $(1, 1, 0)$, $(2, 1, 0)$ and $(1, 1, 1)$. Which agrees with the autoarima results.

```{r}
ar110 = Arima(treasury2, c(1, 1, 0))
ar210 = Arima(treasury2, c(2, 1, 0))
ar111 = Arima(treasury2, c(1, 1, 1))
```

```{r}
{
  checkresiduals(ar110)
  checkresiduals(ar210)
  checkresiduals(ar111)
}
```


# Regression

using treasury as xreg:
```{r}
autofit_with_regression = auto.arima(interest2, xreg = treasury2, approximation = FALSE, stepwise = FALSE, d=0)
summary(autofit_with_regression)
```

plotting a forcast

```{r}
treasury_reg = rep(mean(Treasury[, 2]), 30)
```

```{r}
autoplot(forecast(autofit_with_regression, xreg = treasury_reg))
```

```{r}
checkresiduals(autofit_with_regression)
```

# Fourier extrapolation

```{r}
interest2.k = fft(interest2)

freq = 12*seq_along(interest2)/length(interest2)

t = seq_along(interest2)/12
h = 36
t_forecast = c(1:(length(interest2) + h))/12

get.trajectory <- function(X.k,ts,acq.freq, h) {
  
  N   <- length(ts)
  i   <- complex(real = 0, imaginary = 1)
  x.n <- rep(0,N + h)           # create vector to keep the trajectory
  ks  <- c(0:(length(X.k)-1))
  
  for(n in 0:(N-1 + h)) {       # compute each time point x_n based on freqs X.k
    x.n[n+1] <- sum(X.k * exp(i*2*pi*ks*n/N)) / N
  }
  
  x.n
}

plot(t_forecast, get.trajectory(interest2.k, t, 12, h), type = "l") +
lines(t[1:length(interest2)], interest2, type = "l", col = "red")

```


```{r}
interest2.k = fft(interest2)

freq = 12*seq_along(interest2)/length(interest2)

t = seq_along(interest2)/12
h = 36
t_forecast = c(1:(length(interest2) + h))/12

get.trajectory <- function(X.k,ts,acq.freq, h) {
  
  N   <- length(ts)
  i   <- complex(real = 0, imaginary = 1)
  x.n <- rep(0,N + h)           # create vector to keep the trajectory
  ks  <- c(0:(length(X.k)-1))
  
  for(n in 0:(N-1 + h)) {       # compute each time point x_n based on freqs X.k
    x.n[n+1] <- sum(X.k * exp(i*2*pi*ks*n/N)) / N
  }
  
  x.n
}

plot(t_forecast, Mod(get.trajectory(interest2.k[3:290], t, 12, h)), type = "l", ylim = c(-1, 3.5)) +
lines(t[1:length(interest2)], interest2, type = "l", col = "red")

```


```{r}
checkresiduals(ts(get.trajectory(interest2.k[3:290], t, 12, 0), start = c(2000, 1), frequency = 12) - interest2)
```





# Parameters Stability

### Rolling Window

Throughout the following tests, I will be considering the 

```{r}
parameters =c()
errors = c()

for (i in 1:15){
    parameters <- cbind(parameters, coefficients(Arima(window(interest, frequency = 12, start=c(2000+i, 1), end =c(2005+i, 12)), c(2, 1, 0))))
    errors <- cbind(errors, sqrt(diag(vcov(Arima(window(interest, frequency = 12, start=c(2000+i, 1), end =c(2005+i, 12)), c(2, 1, 0)))))
)
}

t_vals = c(2000:2014)
```

due to the small sample sizes I preferred to keep the default aicc optimization scheme, rather than the aic.

```{r}
{
  plot(t_vals, parameters[1, ], type='b', main="ar1", ylim = c(-0.4, 0.2))+
  arrows(x0=t_vals, y0=parameters[1, ]-errors[1, ], x1 = t_vals, y1=parameters[1, ]+errors[1, ], code=3, angle = 90, length = 0.1)
  plot(t_vals, parameters[2, ], type='b', main="ar2", ylim = c(-0.4, 0.2))+
  arrows(x0=t_vals, y0=parameters[2, ]-errors[2, ], x1 = t_vals, y1=parameters[2, ]+errors[2, ], code=3, angle = 90, length = 0.1)
}
```


There is a noticable jumps between the windows that started at 2009 and the windows that started at 2010. this might be due to the aftermath of the 2008 financial crisis.


```{r}
cbind('ar1' = parameters[1,],
      's.e(ar1)' = errors[1, ],
      'ar2' = parameters[2,],
      's.e(ar2)' = errors[2, ]
      )
```

