---
title: "Report 8"
author: "Khaled Hasan"
date: "2024-05-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(fpp2)
library(tseries)
library(urca)
library(FinTS)
```


## Data
```{r}
GLOBALTEMPERATURE = read.csv(file = "C:\\Users/ss/Desktop/Time_series_Analysis/MyGlobalTemperetures.csv")
global_temp = ts(GLOBALTEMPERATURE[,1], start = c(1850, 1), frequency = 12)
northern_temp = ts(GLOBALTEMPERATURE[,2], start = c(1850, 1), frequency = 12)
southern_temp = ts(GLOBALTEMPERATURE[,3], start = c(1850, 1), frequency = 12)
Regional_temp = read.csv(file = "C:/Users/ss/Desktop/Time_series_Analysis/Regional_temperetures_data.csv")
Europe_temp = ts(Regional_temp[, 3], start = c(1850, 1), frequency = 12)
time = ts(Regional_temp[, 1], start = c(1850, 1), frequency = 12)
```

## Plots

```{r}
library(ggplot2)
library(gridExtra)


autoplot1 <- autoplot(window(global_temp, start = c(2000, 1), freq = 12), ylab = "global")

autoplot2 <- autoplot(window(northern_temp, start = c(2000, 1), freq = 12), ylab = "Northern")

autoplot3 <- autoplot(window(southern_temp, start = c(2000, 1), freq = 12), ylab = "Southern")

par(mfrow = c(1, 3))
plot(autoplot1)
plot(autoplot2)
plot(autoplot3)

layout(matrix(c(1, 2, 3), nrow = 1))
grid.arrange(autoplot1, autoplot2, autoplot3)
```
Notice that the temperature averages in the southern hemi-sphere are out of phase with those from the northern Hemisphere, and that the global average is dominated by the northern hemisphere (northern hemisphere has more land mass).


## plotting p-value

### Finding best fitting model:
```{r}
temp_1980_afterwards = window(global_temp, start=c(1980, 1), freq=12)

global_fitting_sarima = auto.arima(window(global_temp, start=c(1980, 1), freq=12), approximation = FALSE, seasonal = TRUE, stepwise = TRUE, allowdrift = TRUE)

summary(global_fitting_sarima)
```


```{r}
checkresiduals(global_fitting_sarima)
```


```{r}
global_fitting_sarima %>% forecast(h = 24) %>% autoplot()
```


```{r}
minimum = global_fitting_sarima$aic
ord = c(2, 0, 2)
for (i in 0:5) {
  for (j in 0:5) {
    tord = c(i, 0, j)
    if(Arima(temp_1980_afterwards, tord, c(1, 1, 1), include.drift = TRUE)$aic < minimum){
      
      minimum = Arima(temp_1980_afterwards, tord, c(1, 1, 1), include.drift = TRUE)$aic
      ord = tord
    }
  }
}

model = Arima(temp_1980_afterwards, ord, c(1, 1, 1), include.drift = TRUE)
summary(model)
```
Notice that the manually fitted model has a lower aic, a higher loglikelihood and included a drift, that agrees with the trend regression fitting in the following section

```{r}
checkresiduals(model)
```





### Fitting with regression:
```{r}
Arima_fittng <- function(timeseries, startingPoint = start(timeseries), endingPoint = end(timeseries)){
  cutted_data = window(timeseries, start = startingPoint, end = endingPoint, freq = 12)
  t = seq_along(cutted_data)
  regressors = cbind(sin(pi/6*t), cos(pi/6*t), t)
  arima_fit = auto.arima(cutted_data, xreg = regressors, approximation = FALSE, seasonal = TRUE)
  return(arima_fit)
}

global_fitting_Arimareg = Arima_fittng(global_temp, startingPoint = c(1987, 1))

summary(global_fitting_Arimareg)
```


```{r}
checkresiduals(global_fitting_Arimareg)
```

doing the same with the data obtained from Europe
```{r}
Europe_fitting_Arimareg = Arima_fittng(Europe_temp, startingPoint = c(1987, 1))

summary(Europe_fitting_Arimareg)
```

```{r}
checkresiduals(Europe_fitting_Arimareg)
```


## Plotting global warming coefficients

```{r}

linear_coef <- function(DATA, x, Ord, sOrd, radius = 2){
  
  temporary_data = window(DATA, start = c(x-radius, 1), end = c(x+radius, 1))
  
  new_t <- seq_along(temporary_data)
  
  temporary_xreg = cbind(sin(new_t*pi/6), cos(new_t*pi/6), new_t)
  
  temporary_model = arima(temporary_data, order = Ord, seasonal = sOrd, xreg = temporary_xreg)
  
  std_error <- sqrt(diag(vcov(temporary_model)))
  
  return(c(as.numeric(temporary_model$coef["new_t"]), as.numeric(sqrt(diag(vcov(temporary_model)))["new_t"])))
}


plot_Global_warming <- function(timeseries, arima_fit){
  
  #arima_fit = Arima_fittng(timeseries, startingPoint = c(1980, 1))
  
  
  ord = arima_fit$arma
  
  p = ord[1]; q = ord[2]; P = ord[3]; Q = ord[4]; period = ord[5]; d = ord[6]; D = ord[7];
  
  parameters =c()
  
  errors = c()
  
  rad = 20
  
  sp = 1850
  
  fp = 2023
  
  for (i in (sp + rad):(fp - rad)){
      u = linear_coef(timeseries,  i, c(p, d, q), c(P, D, Q) , rad)
      parameters <- cbind(parameters, u[1])
      errors <- cbind(errors, u[2])
  }
  
  t_vals = c((sp + rad):(fp - rad))
  
  MyPlot = (plot(t_vals, parameters, type='b', main=paste("c\nradius: ", toString(rad)), ylim = c(-0.010, 0.010)) +
    arrows(x0=t_vals, y0=parameters-2*errors, x1 = t_vals, y1=parameters+2*errors, code=3, angle = 90, length = 0.1))
  
  return(list(arima_fit, parameters, errors, MyPlot))
}

Results_for_global = plot_Global_warming(global_temp, arima_fit = global_fitting_Arimareg)
```


```{r}
Results_for_Europe = plot_Global_warming(Europe_temp, arima_fit = Europe_fitting_Arimareg)
```

## Plotting p-values

using the standard errors and assuming normal distribution (due to the big number of data points)

```{r}
plot_p_vals <- function(timeseries, arima_fit, r){
  
  #arima_fit = Arima_fittng(timeseries, startingPoint = c(1980, 1))
  
  ord = arima_fit$arma
  
  p = ord[1]; q = ord[2]; P = ord[3]; Q = ord[4]; period = ord[5]; d = ord[6]; D = ord[7];
  
  parameters =c()
  
  errors = c()
  
  p_values = c()
  
  rad = r
  
  sp = 1850
  
  fp = 2023
  
  for (i in (sp + rad):(fp - rad)){
      u = linear_coef(timeseries,  i, c(p, d, q), c(P, D, Q) , rad)
      parameters <- cbind(parameters, u[1])
      errors <- cbind(errors, u[2])
      
      p = 2*pnorm(min(0, 2*u[1]), mean = u[1], sd = u[2], lower.tail = TRUE)
      p_values = cbind(p_values, p)
  }
  
  t_vals = c((sp + rad):(fp - rad))
  
  plot(t_vals, -log10(p_values), type='b', main=paste("p\nradius: ", toString(rad)))
  
}

plot_p_vals(global_temp, arima_fit = global_fitting_Arimareg, 10)
```

```{r}
plot_p_vals(global_temp, arima_fit = global_fitting_Arimareg, 20)
```


```{r}
plot_p_vals(global_temp, arima_fit = global_fitting_Arimareg, 5)
```

```{r}
plot_p_vals(global_temp, arima_fit = global_fitting_Arimareg, 5)
```

